SPP-Perc: Peceptron Learning Based Enhanced Signature Path Prefetcher

### 2019-05-22 back.tex
Minor typo
Cleaning up comments
ran spellcheck

% [EB:] Removing this section title and moving relevant text to next section
%       We weren't talking about LA prefetchers here anyway, just SPP
% \subsection{Lookahead-based Prefetchers}
%% Simple stride prefetching techniques only detect sequences of addresses that differ by a constant value and fail to capture diverse memory access patterns. 
% Lookahead prefetchers can learn complicated patterns by collecting
% histories of observed data access patterns.  EB: Didn't really
% understand this:
%% "and correlating these with the strides, by considering each new page in a vacuum."

@@

% djimenez: it does look like we're pushing it with all the stuff about SPP.
% Let's describe it enough to give the reader the idea that he/she has a vague
% understanding of how it works and can refer to the MICRO paper for more
% details. (I'm not saying cut anything that's already there, just don't expand.)
%% NOTE: Can we do away with this example? 
%% If it looks like we are writing too much about SPP
%% EDIT: Decided to reorganize and remove this for now.
%% 	Consider the case that incoming page number 10 with a block offset 3
%% 	finds a match in the ST. The retrieved pattern signature is 0x30 and
%% 	the Last Offset is 1. Since now there is the Last Block Offset (1),
%% 	Incoming Block Offset (3), Old Signature (0x30), and New Signature
%% 	calculated as per above equation (0x182), SPP can infer
%% 	non-speculatively that the given pattern of memory accesses (as
%% 	captured in the Old Signature) leads to the particular Delta. In
%% 	general, Delta is defined as the difference between the prefetch
%% 	suggestion block and the initial block which triggered the prefetch.
%% 	In this case, since SPP is in learning phase, it is defined as the
%% 	difference between the Incoming Block Offset and the Last Block Offset
%% 	(+2 in this case). This, in turn generates the new memory pattern
%% 	(New Signature). This newly learned signature pattern and the delta
%% 	is stored in the Pattern Table.
%  -- This looks like a detail that is confusing and not necessary to
%  understand SPP:
% \noindent \textbf{Prefetch Table:} The Prefetch Table is a 1024 entry
% direct-mapped table that keeps a record of last few entries
% prefetched.  This proves useful for updating the prefetcher state if a
% tracked prefetch leads to a demand hit or a cache
% eviction. \textit{Note:} The original paper refers to this as the
% Prefetch Filter but since in our context filter refers to the
% perceptron filter, we will be calling this structure as the Prefetch
% Table to avoid any confusion.

@@

% djimenez: this detail is unnecessary and likely to invite questions.
% reviewers might ask "what about linear separability?" but the proof is in the
% pudding.
%This form of indexing makes sure that the hyper-plane learned by the
%perceptron weights is able to differentiate between linearly
%inseparable outcomes.

%\textit{Deviations From Actual Perceptrons:} Traditionally, a perceptron
%prediction involves multiplying the vector of input features:
%F\textsubscript{1xN}, with the corresponding weight vector:
%W\textsubscript{Nx1} in a dot product to obtain the sum y\textsubscript{out}.
%Here what we use is a perceptron-like structure.  The feature is used to hash
%into the weight of perceptrons and the retrieved weights are added straight
%away.  Thus, the perceptron algorithm doesn't introduce any multiplication
%operations in the inference or training process.  Hence what we adapt in this
%work is a perceptron-like learning algorithm as it involves the same principle
%involved in perceptron inferencing and training.

@@

%% Compared to some of the other state of the art prefetchers, SPP is
%% less aggressive.  In a single-core environment, there is no resource
%% contention among the different cores, hence aggressive prefetching is bound to prove beneficial. 
%% As we increase the core
%% count, we observe that SPP starts outperforming rest of the prefetchers.  This
%% can be attributed to the fact that each prefetch suggested by SPP is a
%% carefully calculated and prevents cache pollution.  \textit{<Figure
%% showing the variation of SPP vs BOP wrt core count>} For 4-core applications,
%% SPP suggests XX\% fewer prefetches than Best Offset Prefetcher ~\cite{bop} and yet leads to higher IPC.
%% 
%% The above analysis shows that with a more careful filtering mechanism, any
%% prefetcher like SPP can be tuned to become much more aggressive, yielding higher coverage.  The responsibility of maintaining high accuracy now falls on the
%% independent filter.  To test the hypothesis, we tuned down SPP to the
%% minimum possible threshold, with the effect of increasing prefetch suggestions
%% made by SPP by XX\%.  \textit{<Comparison of SPP-Unleased with SPP / BOP>}
%% Obviously this came at a cost of increased DRAM traffic and cache pollution.

% EB: This remains one of the untouched subsections of the paper

@@

\subsection{Perceptron Learning}
\label{sec:Background-Perceptron}
Perceptron learning for microarchitectural prediction was introduced
for branch prediction~\cite{PerceptronPredictor}. Our predictor uses a
version of microarchitectural perceptron prediction known as the
``hashed perceptron'' organization~\cite{HashedPerceptron}. As an
abstract idea, a hashed perceptron predictor hashes several different
features into values that index several distinct tables. Small integer
weights are read out from the tables and summed. If the sum exceeds
some threshold, a positive prediction is made, {\em e.g.} ``predict
branch taken'' or ``allow the prefetch.'' Otherwise, a negative
prediction is made. Once the ground truth is known, the weights
corredponding to the prediction are incremented if the outcome was
positive, or decremented if it was negative. This update only occurs
if the prediction was incorrect or if the magnitude of the sum failed
to exceed a threshold.  Beyond branch prediction, perceptron learning
has been applied to last-level cache reuse
prediction~\cite{Perc_Reuse,Multiperspective}. In this paper, we apply
it for the first time to prefetch filtering.

%\textit{Perceptron Learning Rule:} In a perceptron-based mechanism, a uniform
%perceptron update principle is followed.  The weights need to be updated if
%the prediction was wrong or the magnitude of the dot product does not exceed a
%certain threshold.  After a certain training period, these weights are
%proportional to the probability of a correct prediction.  Thresholding the
%outcome to a
% magnitude makes sure the perceptrons are trained until a certain level of
%confidence and yet they are not over-trained to the given set.  The perceptron
%learning rule will be applied to PPF.

%ETG: this is what I had added about perceptrons, we're trying to 
%figure out what to leave and what to get rid of ...  
% djimenez: I ended up cutting everything you wrote as well as the figure. I
% wrote something similar above, and the perceptron model described here isn't
% the hashed perceptron that we're using but rather the dot-product model from
% the original paper.

%\subsection{Perceptron-based Prediction}
%
%Perceptrons in microarchitecture were first introduced by Jimenez and Lin~\cite{PerceptronPredictor} 
%for branch prediction. In this work, a perceptron weights vector is selected by a hash of 
%the branch address and its dot product with an input vector of the previous branch history 
%outcomes. If the product is at least 0, the branch is predicted as taken, otherwise it is 
%predicted not taken. The weights are updated with the perceptron update rule: if the prediction 
%is incorrect, or the dot product does not exceed some threshold magnitude, then the weights 
%are incremented or decremented based on their correlation to the corresponding history bits. 
%Perceptron learning yields to superior accuracy, which results in increased opportunity for 
%optimizations.

%\begin{figure}[ht]
%\includegraphics[width=\columnwidth]{Perc_Model}
%\caption{The Basic Perceptron Model}
%\label{Fig:Perc_Model}
%\end{figure}

%\subsubsection{Perceptron Learning}
%
%In a perceptron-based mechanism, a uniform perceptron update principle is
%followed. As shown in the figure~\ref{Fig:Perc_Model} Perceptrons are a simple
%model of neurons in neural networks \cite{perceptrons1,perceptrons2} modeled
%by vectors of signed weights learned through online training. The output of a
%perceptron is the dot product of the weights and a vector of inputs. In this
%work, we make use of the perceptron learning algorithm. There are two
%components to the abstract perceptron learning algorithm:
%
%\begin{enumerate}
%\item To predict true or false, a vector of signed weights is chosen according to some criteria. 
%The dot product of the weights and input vectors, called $y_{\mbox{\small out}}$, is computed. 
%If $y_{\mbox{\small out}}$ exceeds some threshold, the prediction is true, otherwise it is false.
%
%\item To update the weights after the true outcome of the predicted event is
%known, we first consider the value of $y_{\mbox{\small out}}$. If the
%prediction was correct and $|y_{\mbox{\small out}}|$ exceeds a threshold
%$\theta$, then the weights remain unchanged. Otherwise, the inputs are used to
%update the corresponding weights.  If there is positive correlation between
%the input and the outcome of the event, the corresponding weight is
%incremented; otherwise, it is decremented. Over time, the weights are
%proportional to the probability that the outcome of the event is true in the
%context of the input and the criterion for choosing that weight. The weights
%saturate at maximum and minimum values so they will fit in a fixed bit width.
%\end{enumerate}

>>>>

@@

@@

@@

@@

\subsection{Perceptron Learning}
\label{sec:Background-Perceptron}
Perceptron learning for microarchitectural prediction was introduced
for branch prediction~\cite{PerceptronPredictor}. Our predictor uses a
version of microarchitectural perceptron prediction known as the
``hashed perceptron'' organization~\cite{HashedPerceptron}. As an
abstract idea, a hashed perceptron predictor hashes several different
features into values that index several distinct tables. Small integer
weights are read out from the tables and summed. If the sum exceeds
some threshold, a positive prediction is made, {\em e.g.} ``predict
branch taken'' or ``allow the prefetch.'' Otherwise, a negative
prediction is made. Once the ground truth is known, the weights
corresponding to the prediction are incremented if the outcome was
positive, or decremented if it was negative. This update only occurs
if the prediction was incorrect or if the magnitude of the sum failed
to exceed a threshold.  Beyond branch prediction, perceptron learning
has been applied to last-level cache reuse
prediction~\cite{Perc_Reuse,Multiperspective}. In this paper, we apply
it for the first time to do prefetch filtering.

### 2019-05-22 related.tex
ran spellcheck

Unlike spatial prefetchers, lookahead prefetchers take program order
into account when they make predictions.  Shevgoor \textit{et al.}
propose the Variable Length Delta Prefetcher (VLDP)~\cite{VLDP}, which
correlates histories of deltas between cache line accesses within
memory pages with the next delta within that page. SPP~\cite{SPP} and
KPC's prefetching component~\cite{KPC} are more recent examples of
lookahead prefetchers. They try to predict not only what data will be
used in the future, but also the precise order in which the data will
be used, within a given page. Predictions made by lookahead
prefetchers can be fed back into their prediction mechanisms to
predict even further down a speculative path of memory acesses. These
prefetchers can also generalize their learned patterns from one page,
and use those patterns to make predictions in other pages.

@@

A low-accuracy aggressive prefetcher can significantly harm
performance.  To minimize interference from prefetching, Wu \textit{et
  al.} propose PACMan~\cite{pacman}, a prefetch-aware cache management
policy. PACMan dedicates some LLC sets to each of three competing
policies that treat demand and prefetch requests differently, using
the policy in the rest of the cache that shows the fewest
misses. Seshadri \textit{et al.} propose ICP~\cite{icp}, which demotes
a prefetch to the lowest reuse priority on a demand hit, based on the
observation that most prefetches are dead after their first hit. To
address prefetcher-caused cache pollution, it also uses a variation of
EAF~\cite{eaf} to track prefetching accuracy, and inserts only
accurate prefetches to the higher priority position in the LRU
stack. Jain \textit{et al.} propose Harmony~\cite{Harmony} to
accomodate prefetches in their MIN algorithm-inspired Hawkeye cache
management system. Ebrahimi \textit{et al.} introduce
HPAC~\cite{HPAC} which provides a coordinated control between 
multiple prefetchers present in a CMP by looking at the 
prefetcher-induced inter-core interferance.

@@

In addition to branch prediction~\cite{PerceptronPredictor},
perceptron-based learning has been applied to the area of cache
management.  Teran \textit{et al.} propose using perceptrons to
predict cache line resuse, bypass, and replacement~\cite{Perc_Reuse}.
Perceptron Learning trains weights selected by hashes of multiple
features, including the PC of the memory access instruction, some
other recent PCs, and two different shifts of the tag of the
referenced block. These features are used to index into weight tables,
and the weights are then thresholded to generate a prediction. When a
block from one of a few sampled sets~\cite{sdbp} is reused or evicted,
the corresponding weights are decremented or incremented, according to
the perceptron learning rule. Multiperspective Reuse
Prediction~\cite{Multiperspective} improves on Perceptron Learning by
contributing many new features.

>>>>

Unlike spatial prefetchers, lookahead prefetchers take program order
into account when they make predictions.  Shevgoor \textit{et al.}
propose the Variable Length Delta Prefetcher (VLDP)~\cite{VLDP}, which
correlates histories of deltas between cache line accesses within
memory pages with the next delta within that page. SPP~\cite{SPP} and
KPC's prefetching component~\cite{KPC} are more recent examples of
lookahead prefetchers. They try to predict not only what data will be
used in the future, but also the precise order in which the data will
be used, within a given page. Predictions made by lookahead
prefetchers can be fed back into their prediction mechanisms to
predict even further down a speculative path of memory accesses. These
prefetchers can also generalize their learned patterns from one page,
and use those patterns to make predictions in other pages.

@@

A low-accuracy aggressive prefetcher can significantly harm
performance.  To minimize interference from prefetching, Wu \textit{et
  al.} propose PACMan~\cite{pacman}, a prefetch-aware cache management
policy. PACMan dedicates some LLC sets to each of three competing
policies that treat demand and prefetch requests differently, using
the policy in the rest of the cache that shows the fewest
misses. Seshadri \textit{et al.} propose ICP~\cite{icp}, which demotes
a prefetch to the lowest reuse priority on a demand hit, based on the
observation that most prefetches are dead after their first hit. To
address prefetcher-caused cache pollution, it also uses a variation of
EAF~\cite{eaf} to track prefetching accuracy, and inserts only
accurate prefetches to the higher priority position in the LRU
stack. Jain \textit{et al.} propose Harmony~\cite{Harmony} to
accommodate prefetches in their MIN algorithm-inspired Hawkeye cache
management system. Ebrahimi \textit{et al.} introduce
HPAC~\cite{HPAC} which provides a coordinated control between 
multiple prefetchers present in a CMP by looking at the 
prefetcher-induced inter-core interference.

@@

In addition to branch prediction~\cite{PerceptronPredictor},
perceptron-based learning has been applied to the area of cache
management.  Teran \textit{et al.} propose using perceptrons to
predict cache line reuse, bypass, and replacement~\cite{Perc_Reuse}.
Perceptron Learning trains weights selected by hashes of multiple
features, including the PC of the memory access instruction, some
other recent PCs, and two different shifts of the tag of the
referenced block. These features are used to index into weight tables,
and the weights are then thresholded to generate a prediction. When a
block from one of a few sampled sets~\cite{sdbp} is reused or evicted,
the corresponding weights are decremented or incremented, according to
the perceptron learning rule. Multiperspective Reuse
Prediction~\cite{Multiperspective} improves on Perceptron Learning by
contributing many new features.
